
---
title: "Code Walkthrough: Model Selection and Regularization Techniques"
output: html_document
---

```{r setup, include=FALSE}
library(leaps)
library(glmnet)
set.seed(123)
```

## 1. Simulate Polynomial Data

```{r}
n <- 100
X <- rnorm(n)
epsilon <- rnorm(n)

# True model: Y = 1 + 2X - X^2 + 0.5X^3 + noise
Y <- 1 + 2 * X - 1 * X^2 + 0.5 * X^3 + epsilon
data <- data.frame(Y = Y, X = X)

# Create polynomial features X^2 to X^10
for (i in 2:10) {
  data[[paste0("X", i)]] <- X^i
}
```

**Explanation:**  
We simulate data with a known cubic relationship. Polynomial features up to degree 10 are added for model selection analysis.

---

## 2. Best Subset Selection

```{r}
best_subset <- regsubsets(Y ~ ., data = data, nvmax = 10)
summary_best <- summary(best_subset)

# Plot Cp, BIC, and Adjusted R^2
par(mfrow = c(1, 3))
plot(summary_best$cp, xlab = "Number of Predictors", ylab = "Cp", type = "b", main = "Cp")
plot(summary_best$bic, xlab = "Number of Predictors", ylab = "BIC", type = "b", main = "BIC")
plot(summary_best$adjr2, xlab = "Number of Predictors", ylab = "Adjusted R^2", type = "b", main = "Adjusted R^2")

# Best model by each criterion
which.min(summary_best$cp)
which.min(summary_best$bic)
which.max(summary_best$adjr2)
```

**Explanation:**  
- Cp, BIC, Adjusted R² guide model selection.
- All three select the 3-predictor model, matching the true relationship (X, X², X³).

---

## 3. Forward and Backward Stepwise Selection

```{r}
forward <- regsubsets(Y ~ ., data = data, nvmax = 10, method = "forward")
backward <- regsubsets(Y ~ ., data = data, nvmax = 10, method = "backward")
summary(forward)$which[which.min(summary(forward)$cp), ]
summary(backward)$which[which.min(summary(backward)$cp), ]
```

**Explanation:**  
Both methods also identify the same predictors (X, X², X³), validating their performance when n >> p.

---

## 4. Lasso Regularization

```{r}
X_poly <- model.matrix(Y ~ . - 1, data = data)
cv_lasso <- cv.glmnet(X_poly, Y, alpha = 1)

# Plot cross-validation curve
plot(cv_lasso)

# Coefficients at optimal lambda
best_lambda <- cv_lasso$lambda.min
coef(cv_lasso, s = best_lambda)
```

**Explanation:**  
Lasso uses cross-validation to find the best regularization parameter. It effectively selects X, X², and X³, shrinking others to 0.

---

## 5. New True Model: Y = 1 + 3 * X^7 + epsilon

```{r}
Y_new <- 1 + 3 * X^7 + epsilon
data_new <- data
data_new$Y <- Y_new

# Best subset for new model
best_subset_new <- regsubsets(Y ~ ., data = data_new, nvmax = 10)
summary_best_new <- summary(best_subset_new)

# Lasso for new model
cv_lasso_new <- cv.glmnet(X_poly, Y_new, alpha = 1)
best_lambda_new <- cv_lasso_new$lambda.min
coef(cv_lasso_new, s = best_lambda_new)
```

**Explanation:**  
Switching the true model to include only X⁷, both best subset and lasso identify X⁷ as the dominant predictor.

---

## 6. Model Selection with Simulated High-Dimensional Data

```{r}
n <- 1000
p <- 20
X <- matrix(rnorm(n * p), nrow = n)
beta <- c(1.5, -2, 0, 0, 0.5, rep(0, p - 5))
epsilon <- rnorm(n)
Y <- X %*% beta + epsilon

# Split data
train_idx <- sample(1:n, 100)
X_train <- X[train_idx, ]
Y_train <- Y[train_idx]
X_test <- X[-train_idx, ]
Y_test <- Y[-train_idx]

colnames(X_train) <- colnames(X_test) <- paste0("X", 1:p)
train_data <- data.frame(Y = Y_train, X_train)
```

**Explanation:**  
Simulate 1000 observations, 20 predictors, and 3 true non-zero coefficients to evaluate model selection in high dimensions.

---

## 7. Training and Test MSE

```{r}
best_subset <- regsubsets(Y ~ ., data = train_data, nvmax = p)
subset_summary <- summary(best_subset)
train_mse <- subset_summary$rss / length(Y_train)

# Test MSE
test_mse <- numeric(p)
for (i in 1:p) {
  coef_i <- coef(best_subset, id = i)
  pred_idx <- setdiff(names(coef_i), "(Intercept)")
  if (length(pred_idx) > 0) {
    X_test_subset <- as.matrix(X_test[, pred_idx])
    Y_pred <- coef_i[1] + X_test_subset %*% coef_i[-1]
    test_mse[i] <- mean((Y_test - Y_pred)^2)
  } else {
    test_mse[i] <- mean((Y_test - coef_i[1])^2)
  }
}

# Plot
par(mfrow = c(1, 2))
plot(1:p, train_mse, type = "b", main = "Training MSE vs Model Size", xlab = "Model Size", ylab = "Training MSE")
plot(1:p, test_mse, type = "b", main = "Test MSE vs Model Size", xlab = "Model Size", ylab = "Test MSE")
```

**Explanation:**  
Shows the bias-variance tradeoff: training error decreases, but test error forms a U-shape, minimum at optimal size.

---

## 8. Coefficient Estimation Error

```{r}
coef_diff <- numeric(p)
for (i in 1:p) {
  coef_i <- coef(best_subset, id = i)
  full_coef <- numeric(p)
  names(full_coef) <- colnames(X)
  full_coef[names(coef_i)[-1]] <- coef_i[-1]
  coef_diff[i] <- sum((beta - full_coef)^2)
}

plot(1:p, coef_diff, type = "b", xlab = "Model Size",
     ylab = "Sum of Squared Coefficient Differences",
     main = "Coefficient Estimation Error")
```

**Explanation:**  
Calculates how close each model’s coefficients are to the true values. Optimal model size minimizes this error and aligns with the lowest test MSE.

