
---
title: "Code Walkthrough: Decision Trees, Bagging, Random Forest, and SVM Classification"
output: html_document
---

```{r setup, include=FALSE}
library(tree)
library(randomForest)
library(e1071)
set.seed(42)
```

## Part 1: Regression Trees and Ensemble Methods

### Load and Explore Data

```{r}
carseats <- read.csv("Carseats.csv")
str(carseats)
```

### Data Split

```{r}
train_indices <- sample(1:nrow(carseats), 0.7 * nrow(carseats))
train <- carseats[train_indices, ]
test <- carseats[-train_indices, ]
```

### Fit a Regression Tree

```{r}
reg_tree <- tree(Sales ~ ., data = train)
plot(reg_tree)
text(reg_tree, pretty = 0)
```

**Explanation:**  
A regression tree is fitted to predict `Sales`. The tree is visualized to understand decision splits.

### Evaluate Tree Performance

```{r}
test_predictions <- predict(reg_tree, newdata = test)
test_mse <- mean((test$Sales - test_predictions)^2)
test_mse
```

**Explanation:**  
The test MSE indicates how well the tree generalizes to new data.

### Cross-Validation and Pruning

```{r}
cv_tree <- cv.tree(reg_tree)
plot(cv_tree$size, cv_tree$dev, type = "b", xlab = "Tree Size", ylab = "CV Error")

optimal_size <- which.min(cv_tree$dev)
pruned_tree <- prune.tree(reg_tree, best = cv_tree$size[optimal_size])
plot(pruned_tree)
text(pruned_tree, pretty = 0)

pruned_predictions <- predict(pruned_tree, newdata = test)
pruned_mse <- mean((test$Sales - pruned_predictions)^2)
pruned_mse
```

**Explanation:**  
Cross-validation helps identify the optimal tree size. Pruning simplifies the model and may improve generalization.

### Bagging Model

```{r}
bagging_model <- randomForest(Sales ~ ., data = train, mtry = ncol(train) - 1, importance = TRUE)
bagging_predictions <- predict(bagging_model, newdata = test)
bagging_mse <- mean((test$Sales - bagging_predictions)^2)
bagging_mse
varImpPlot(bagging_model)
```

**Explanation:**  
Bagging averages multiple trees to reduce variance and boost accuracy. The importance plot shows key predictors.

### Random Forest Model

```{r}
random_forest_model <- randomForest(Sales ~ ., data = train, mtry = sqrt(ncol(train) - 1), importance = TRUE)
rf_predictions <- predict(random_forest_model, newdata = test)
rf_mse <- mean((test$Sales - rf_predictions)^2)
rf_mse
varImpPlot(random_forest_model)
```

**Explanation:**  
Random forest improves on bagging by reducing feature correlation. We observe performance and feature importance.

---

## Part 2: Support Vector Machines (SVM)

### Load and Prepare Data

```{r}
oj <- read.csv("OJ.csv")
train_indices <- sample(1:nrow(oj), 800)
train <- oj[train_indices, ]
test <- oj[-train_indices, ]
train$Purchase <- as.factor(train$Purchase)
test$Purchase <- as.factor(test$Purchase)
```

### Linear SVM

```{r}
svc_model <- svm(Purchase ~ ., data = train, kernel = "linear", cost = 0.01)
train_pred <- predict(svc_model, train)
test_pred <- predict(svc_model, test)
train_error <- mean(train_pred != train$Purchase)
test_error <- mean(test_pred != test$Purchase)
train_error
test_error
```

**Explanation:**  
A linear SVM is trained and evaluated. We compute misclassification errors on both sets.

### Tuning Linear SVM

```{r}
tune_result <- tune(svm, Purchase ~ ., data = train, kernel = "linear",
                    ranges = list(cost = seq(0.01, 10, length.out = 10)))
best_cost <- tune_result$best.parameters$cost
optimal_svc_model <- svm(Purchase ~ ., data = train, kernel = "linear", cost = best_cost)
optimal_test_pred <- predict(optimal_svc_model, test)
mean(optimal_test_pred != test$Purchase)
```

**Explanation:**  
Cost parameter tuning improves the linear SVM’s performance by controlling margin flexibility.

### Radial SVM

```{r}
tune_radial <- tune(svm, Purchase ~ ., data = train, kernel = "radial",
                    ranges = list(cost = seq(0.01, 10, length.out = 10)))
best_cost_radial <- tune_radial$best.parameters$cost
radial_model <- svm(Purchase ~ ., data = train, kernel = "radial", cost = best_cost_radial)
mean(predict(radial_model, test) != test$Purchase)
```

**Explanation:**  
Radial kernel allows non-linear decision boundaries. Tuning cost improves classification accuracy.

### Polynomial SVM

```{r}
tune_poly <- tune(svm, Purchase ~ ., data = train, kernel = "polynomial", degree = 2,
                  ranges = list(cost = seq(0.01, 10, length.out = 10)))
best_cost_poly <- tune_poly$best.parameters$cost
poly_model <- svm(Purchase ~ ., data = train, kernel = "polynomial", degree = 2, cost = best_cost_poly)
mean(predict(poly_model, test) != test$Purchase)
```

**Explanation:**  
Polynomial kernel SVM captures interaction effects. Tuning selects the optimal cost for performance.

---

## Comparison of Models

```{r}
error_rates <- data.frame(
  Method = c("Linear SVM", "Radial SVM", "Polynomial SVM"),
  Test_Error = c(mean(optimal_test_pred != test$Purchase),
                 mean(predict(radial_model, test) != test$Purchase),
                 mean(predict(poly_model, test) != test$Purchase))
)
print(error_rates)
```

**Conclusion:**  
Among all SVM models, the radial kernel achieved the lowest test error, indicating it handled non-linear boundaries best.
