
---
title: "Bias-Variance Tradeoff & Auto Dataset Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Question 1 – Conceptual Analysis

- **(a)** Flexible methods preferred with large n, small p (low overfitting risk).
- **(b)** Inflexible methods preferred with small n, large p (high variance).
- **(c)** Flexible methods capture non-linear relationships better.
- **(d)** Inflexible models handle high error variance better by ignoring noise.

---

## Question 2 – Bias-Variance Tradeoff Simulation

### Generate Error Components
```{r bias-variance-sim}
library(ggplot2)
library(reshape2)

flexibility <- seq(1, 15, by = 0.1)
bias_squared <- 1 / flexibility
variance <- log(flexibility) / 3
training_error <- 1 / (flexibility + 1)
test_error <- 0.2 / (flexibility + 1) + variance
bayes_error <- rep(0.1, length(flexibility))

error_data <- data.frame(
  Flexibility = flexibility,
  BiasSquared = bias_squared,
  Variance = variance,
  TrainingError = training_error,
  TestError = test_error,
  BayesError = bayes_error
)

error_data_melted <- melt(error_data, id = "Flexibility")
```

### Plot Bias-Variance Tradeoff
```{r bias-variance-plot}
ggplot(error_data_melted, aes(x = Flexibility, y = value, color = variable, linetype = variable)) +
  geom_line(linewidth = 1) +
  labs(title = "Bias-Variance Tradeoff", x = "Model Flexibility", y = "Error / Value") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  ylim(0, 1)
```

---

## Question 3 – Auto Dataset Analysis

### Load and Clean Data
```{r load-data}
auto_data <- read.table("Auto.data", header = TRUE, na.strings = "?")
auto_data <- na.omit(auto_data)
```

### Identify Predictor Types
```{r predictor-types}
quantitative <- sapply(auto_data, is.numeric)
qualitative <- !quantitative

cat("Quantitative Predictors:", names(auto_data)[quantitative], "\n")
cat("Qualitative Predictors:", names(auto_data)[qualitative], "\n")

quantitative_cols <- names(auto_data)[quantitative]
```

### Range of Quantitative Predictors
```{r range-values}
range_values <- sapply(auto_data[quantitative_cols], function(x) range(x, na.rm = TRUE))
range_values
```

### Mean and Standard Deviation
```{r mean-sd}
mean_values <- sapply(auto_data[quantitative_cols], function(x) mean(x, na.rm = TRUE))
sd_values <- sapply(auto_data[quantitative_cols], function(x) sd(x, na.rm = TRUE))

mean_values
sd_values
```

### Subset Analysis: Remove Observations 10–85
```{r subset-analysis}
auto_data_subset <- auto_data[-(10:85), ]

subset_range <- sapply(auto_data_subset[quantitative_cols], function(x) range(x, na.rm = TRUE))
subset_mean <- sapply(auto_data_subset[quantitative_cols], function(x) mean(x, na.rm = TRUE))
subset_sd <- sapply(auto_data_subset[quantitative_cols], function(x) sd(x, na.rm = TRUE))

subset_range
subset_mean
subset_sd
```

### Scatterplot Matrix for Quantitative Predictors
```{r scatterplot-matrix, fig.width=10, fig.height=10}
library(GGally)
ggpairs(auto_data[quantitative_cols])
```

### Individual Scatterplots: mpg vs Other Predictors
```{r scatterplots}
library(ggplot2)

for (var in quantitative_cols[quantitative_cols != "mpg"]) {
  print(
    ggplot(auto_data, aes_string(x = var, y = "mpg")) +
      geom_point() +
      labs(title = paste("MPG vs", var), x = var, y = "MPG") +
      theme_minimal()
  )
}
```

---

## Conclusion

- Strong negative correlation found between **mpg** and **weight**, **horsepower**, and **displacement**.
- **Year** positively correlates with **mpg**, reflecting advances in fuel efficiency.
- Several predictors are well-suited for modeling **mpg** using regression or other supervised learning techniques.
