
---
title: "Code Walkthrough: Exploring Non-Linearity in Regression"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.keep = "all", fig.path = "figure/")
library(ggplot2)
library(boot)
library(splines)
```

# Question 2: Function Complexity

## (a) Constant Function (λ = ∞, m = 0)

```{r}
plot(1:10, rep(mean(1:10), 10), col = "blue", type = "l", 
     main = "Constant Function", ylab = "y", xlab = "x")
```

**Explanation:**  
A constant function fits all points to the mean, representing an extremely simple model.

## (b) Linear Function (λ = ∞, m = 1)

```{r}
plot(1:10, 1:10, type = "l", col = "blue", main = "Linear Function", ylab = "y", xlab = "x")
```

**Explanation:**  
A linear model fits data with a straight line — a moderate level of flexibility.

## (c) Quadratic Function (λ = ∞, m = 2)

```{r}
x <- seq(-10, 10, length.out = 100)
y <- x^2
plot(x, y, type = "l", col = "red", main = "Quadratic Function", ylab = "y", xlab = "x")
```

**Explanation:**  
Quadratic functions allow modeling of curvature, increasing flexibility over linear models.

## (d) Cubic Function (λ = ∞, m = 3)

```{r}
y <- x^3
plot(x, y, type = "l", col = "purple", main = "Cubic Function", ylab = "y", xlab = "x")
```

**Explanation:**  
Cubic models can model inflection points and are more flexible than quadratics.

## (e) Interpolated Fit (λ = 0, m = 3)

```{r}
x <- 1:10
y <- cos(x) + rnorm(10, sd = 0.5)
plot(x, y, type = "o", col = "orange", main = "Interpolated Fit", ylab = "y", xlab = "x")
```

**Explanation:**  
This demonstrates overfitting, where the model fits all points exactly (zero regularization).

---

# Question 3: Basis Function Regression

```{r}
curve_fit <- function(X) {
  if (X < 1) {
    return(1 + X)
  } else {
    return(1 + X - 2 * (X - 1)^2)
  }
}

X <- seq(-2, 2, by = 0.01)
Y <- sapply(X, curve_fit)

plot(X, Y, type = "l", col = "blue", lwd = 2,
     main = "Basis Function Regression", xlab = "X", ylab = "Y")
abline(v = 1, col = "red", lty = 2)
points(c(-2, 0, 1, 2), sapply(c(-2, 0, 1, 2), curve_fit), pch = 16)
```

**Explanation:**  
We build a piecewise function that models the data using a custom basis. This shows how basis expansions can capture non-linear behavior.

---

# Question 9: Modeling NOX from DIS (Boston Dataset)

```{r}
data <- read.table("Boston.data", header = FALSE)
colnames(data) <- c("CRIM", "ZN", "INDUS", "CHAS", "NOX", "RM", "AGE", 
                    "DIS", "RAD", "TAX", "PTRATIO", "B", "LSTAT")
```

## (a) Cubic Polynomial Regression

```{r}
fit_cubic <- lm(NOX ~ poly(DIS, 3), data = data)
summary(fit_cubic)

ggplot(data, aes(x = DIS, y = NOX)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "lm", formula = y ~ poly(x, 3), color = "red", se = FALSE) +
  labs(title = "Cubic Polynomial Fit", x = "DIS", y = "NOX")
```

**Explanation:**  
Fits a cubic polynomial to model `NOX` using `DIS`, capturing non-linear trends.

## (b) Polynomial Fits for Degrees 1 to 10

```{r}
rss_values <- numeric()
for (degree in 1:10) {
  fit <- lm(NOX ~ poly(DIS, degree), data = data)
  rss_values[degree] <- sum(residuals(fit)^2)
}
plot(1:10, rss_values, type = "b", col = "blue", 
     xlab = "Degree", ylab = "RSS", main = "RSS for Polynomial Fits")
```

**Explanation:**  
RSS decreases as degree increases, but this may lead to overfitting.

## (c) Cross-Validation for Polynomial Degree

```{r}
cv_error <- numeric()
for (degree in 1:10) {
  fit <- glm(NOX ~ poly(DIS, degree), data = data)
  cv_error[degree] <- cv.glm(data, fit)$delta[1]
}
plot(1:10, cv_error, type = "b", col = "red", pch = 19,
     xlab = "Polynomial Degree", ylab = "CV Error",
     main = "Cross-Validation for Polynomial Degrees")
```

**Explanation:**  
Selects the degree with lowest CV error, balancing bias and variance.

## (d) Regression Spline with df = 4

```{r}
fit_spline <- lm(NOX ~ bs(DIS, df = 4), data = data)
summary(fit_spline)

ggplot(data, aes(x = DIS, y = NOX)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "lm", formula = y ~ bs(x, df = 4), color = "red", se = FALSE) +
  labs(title = "Regression Spline Fit (df = 4)", x = "DIS", y = "NOX")
```

**Explanation:**  
Regression splines allow flexible fitting using piecewise polynomials with smooth transitions.

## (e) RSS for Splines with Varying df

```{r}
rss_spline <- numeric()
for (df in 3:10) {
  fit <- lm(NOX ~ bs(DIS, df = df), data = data)
  rss_spline[df] <- sum(residuals(fit)^2)
}
plot(3:10, rss_spline[3:10], type = "b", col = "blue", 
     xlab = "Degrees of Freedom", ylab = "RSS",
     main = "RSS for Regression Splines")
```

**Explanation:**  
Evaluates spline flexibility more degrees of freedom allow better fit but may overfit.

## (f) Cross-Validation for Splines

```{r}
cv_error_spline <- numeric()
for (df in 3:10) {
  fit <- glm(NOX ~ bs(DIS, df = df), data = data)
  cv_error_spline[df] <- cv.glm(data, fit)$delta[1]
}
plot(3:10, cv_error_spline[3:10], type = "b", col = "red", pch = 19,
     xlab = "Degrees of Freedom", ylab = "CV Error",
     main = "CV Error for Regression Splines")
```

**Explanation:**  
Cross-validation identifies the optimal degree of freedom for spline smoothing.
