
---
title: "RSM8512_Assignment5_1006759189_Non-Linearity"
output:
  pdf_document: default
  html_notebook: default
---

# Question 2
```{r}
knitr::opts_chunk$set(cache = FALSE, fig.keep = "all")
knitr::opts_chunk$set(fig.path = "figure/")


```


## (a) $\lambda = \infty$, m = 0 (Constant Function)
```{r}
plot(1:10, rep(mean(1:10), 10),col = "blue", type="l", main="(Constant Function)")
```

## (b) $\lambda = \infty$, m = 1 (Linear Function)
```{r}
plot(1:10, 1:10, type="l", col = "blue" ,main="(Linear Function)")
```

## (c) $\lambda = \infty$, m = 2 (Quadratic Function)
```{r}
x <- seq(-10, 10, length.out = 100)
y <- x^2
plot(x, y, type="l", col = "red", main="(Quadratic Function)")
```

## (d) $\lambda = \infty$, m = 3 (Cubic Function)
```{r}
y <- x^3
plot(x, y, type="l", col = "purple", main="(Cubic Function)")
```

## (e) $\lambda = 0, m = 3 (Interpolates All Points)
```{r}
x <- 1:10
y <- cos(x) + rnorm(10, sd = 0.5)
plot(x, y, type="o",col = "orange", main="(Interpolates All Points)")
```

# Question 3


```{r}
# Basis Functions for Regression
curve_fit <- function(X) {
  if (X < 1) {
    return(1 + X) 
  } else {
    return(1 + X - 2 * (X - 1)^2) 
  }
}


X <- seq(-2, 2, by = 0.01)
Y <- sapply(X, curve_fit)


plot(X, Y, type = "l", col = "blue", lwd = 2, xlab = "X", ylab = "Y",
     main = "Linear Regression with Basis Functions")
abline(v = 1, col = "red", lty = 2)  
points(c(-2, 0, 1, 2), sapply(c(-2, 0, 1, 2), curve_fit), col = "black", pch = 16)


text(-2, curve_fit(-2), labels = round(curve_fit(-2), 2), pos = 3)
text(0, curve_fit(0), labels = round(curve_fit(0), 2), pos = 3)
text(1, curve_fit(1), labels = round(curve_fit(1), 2), pos = 3)
text(2, curve_fit(2), labels = round(curve_fit(2), 2), pos = 3)

```


# Question 9


```{r}


data <- read.table("C:/Users/dokan/Downloads/Assignment5/Boston.data", header = FALSE)
colnames(data) <- c("CRIM", "ZN", "INDUS", "CHAS", "NOX", "RM", "AGE", 
                    "DIS", "RAD", "TAX", "PTRATIO", "B", "LSTAT")



```

## (a) Polynomial Regression (Degree 3)
```{r}
fit_cubic <- lm(NOX ~ poly(DIS, 3), data = data)
summary(fit_cubic)

library(ggplot2)
ggplot(data, aes(x = DIS, y = NOX)) +
  geom_point(color = "blue", alpha = 0.5) +
  stat_smooth(method = "lm", formula = y ~ poly(x, 3), color = "red", se = FALSE) +
  labs(title = "Cubic Polynomial Fit for NOX vs DIS", x = "DIS", y = "NOX")
```

## (b) Polynomial Fits for Degrees 1 to 10
```{r}
rss_values <- numeric()


for (degree in 1:10) {
  fit <- lm(NOX ~ poly(DIS, degree), data = data)
  rss <- sum(residuals(fit)^2)
  rss_values <- c(rss_values, rss)
  
  
  ggplot(data, aes(x = DIS, y = NOX)) +
    geom_point(color = "blue", alpha = 0.5) +
    stat_smooth(method = "lm", formula = y ~ poly(x, degree), color = "red", se = FALSE) +
    labs(title = paste("Polynomial Degree", degree, "Fit for NOX vs DIS"),
         x = "DIS", y = "NOX") +
    theme_minimal()
}


rss_values

```

## (c) Cross-Validation for Optimal Polynomial Degree
```{r}
library(boot)


cv_error <- numeric()
for (degree in 1:10) {
  fit <- glm(NOX ~ poly(DIS, degree), data = data)
  cv_error[degree] <- cv.glm(data, fit)$delta[1]
}


plot(1:10, cv_error, type = "b", col = "red", pch = 19, lwd = 2,
     xlab = "Polynomial Degree", ylab = "Cross-Validation Error",
     main = "Cross-Validation Error for Polynomial Degrees")

```

## (d) Regression Spline (4 Degrees of Freedom)
```{r}
library(splines)


fit_spline <- lm(NOX ~ bs(DIS, df = 4), data = data)
summary(fit_spline)


ggplot(data, aes(x = DIS, y = NOX)) +
  geom_point(color = "blue", alpha = 0.5) +
  stat_smooth(method = "lm", formula = y ~ bs(x, df = 4), color = "red", se = FALSE) +
  labs(title = "Regression Spline Fit for NOX vs DIS", x = "DIS", y = "NOX")

```

## (e) Regression Splines for Various Degrees of Freedom
```{r}
rss_spline <- numeric()


for (df in 3:10) {
  fit <- lm(NOX ~ bs(DIS, df = df), data = data)
  rss <- sum(residuals(fit)^2)
  rss_spline <- c(rss_spline, rss)
  
 
  ggplot(data, aes(x = DIS, y = NOX)) +
    geom_point(color = "blue", alpha = 0.5) +
    stat_smooth(method = "lm", formula = y ~ bs(x, df = df), color = "red", se = FALSE) +
    labs(title = paste("Spline Fit with", df, "DF"), x = "DIS", y = "NOX") +
    theme_minimal()
}


rss_spline

```

## (f) Cross-Validation for Optimal Degrees of Freedom
```{r}
cv_error_spline <- numeric()


for (df in 3:10) {
  fit <- glm(NOX ~ bs(DIS, df = df), data = data)
  cv_error_spline[df] <- cv.glm(data, fit)$delta[1]
}


plot(3:10, cv_error_spline[3:10], type = "b", col = "red", pch = 19, lwd = 2,
     xlab = "Degrees of Freedom", ylab = "Cross-Validation Error",
     main = "Cross-Validation Error for Splines")

```
